---
name: Joe Wicke
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
---

Your editor wants to know what a "typical" month of car thefts looks
like in Prince George's County. Not just the average --- she wants to
know how confident you can be in that number. Using monthly figures,
confidence intervals and margin of error, you'll come up with an answer
and decide whether the data supports a clear story. As with the previous
homework, provide answers in this notebook when prompted and you will
have to fill in values where you see REPLACE_ME. Get started by loading
the tidyverse in a new R block.

### Task 1: Load the tidyverse

`{r setup, include=FALSE} knitr::opts_chunk$set(echo = TRUE) # LOAD THE TIDYVERSE`

```{r}
library(tidyverse)
```

### Task 2: Load the Prince George's county crime data

``` {r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)
```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any
trend over time. What's your initial reaction to the change in car
thefts? Is there a trend? If your editor asked you "how many car thefts
happen in a typical month?", what would you say based on this chart
alone?

Generally, car thefts have gone down in recent years, but still fluctuate and rise every few months. Based on this, I'd say it's likely to go up soon, but then will plummet again. Because the numbers fluctuate so much, however, I wouldn't be comfortable giving an estimate of how many happen in a typical month. 

``` {r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )
```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the
uncertainty around our average. We'll calculate the standard error
(which measures how much the sample mean could vary), and the margin of
error (which gives us the plus/minus range). Replace the REPLACE_ME
values based on what we did in class.

``` {r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(32),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary
```

**Answer this question:** Explain in your own words what the margin of
error tells us here. If you were writing a story, how would you use this
number? For example, how would you complete this sentence: "Prince
George's County averages about \_\_\_ car thefts per month, plus or
minus \_\_\_."

The margin of error is essentially telling us "Prince George's County averages about 375 car thefts per monnth, plus or minus 34. This would be used in a story to say that the "true" average could be 34 thefts higher or 34 thefts lower. 

### Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval ---
the range where we're 95% confident the true average falls. Replace the
REPLACE_ME values.

``` {r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)
```

**Answer this question:** If a county official claimed "car thefts are
down to about 300 per month," does your confidence interval support that
claim? What about a claim of 350 per month? Explain how the confidence
interval helps you evaluate each claim.

It does not because, at most, the "true" number of average car thefts is 358. A claim of 350 would be a bit of a stretch, but the confidence interval somewhat supports it. However, it would not be the most accurate assessment. 

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence
interval. Does this change anything about your answer from Task 3?

This does change my answer. Given a low confidence interval, I would be very comfortable telling my editor that around 375 car thefts happen per month. 

``` {r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()
```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for
polls. That same principle applies here: more data means a smaller
margin of error. Let's see what would happen if we had 40 months of data
instead of the number we have, assuming the same mean and standard
deviation. Replace the REPLACE_ME value with 40.

``` {r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual data", "40 months"),
  months = c(theft_summary$months, hypothetical_summary$months),
  margin_error = c(theft_summary$margin_error, hypothetical_summary$margin_error),
  ci_lower = c(theft_summary$ci_lower, hypothetical_summary$ci_lower),
  ci_upper = c(theft_summary$ci_upper, hypothetical_summary$ci_upper)
)

comparison
```

**Answer this question:** How did the margin of error and confidence
interval change? What does this tell you about the tradeoff between
having more data and being able to make precise claims? If you were
covering crime trends, how many months of data would you want before
writing a definitive story?

The 40-month data has a smaller confidence interval, which means there's a more precise range of what the "true" mean is. Before writing a definitive story, I think around 5 years (or 60 months) of data should be sufficient. 

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1.  Write a headline and a one-paragraph lede about car thefts in Prince
    George's County. Your text should properly reflect the uncertainty
    in the data --- don't overstate or understate what the numbers show.
    
Headline: Prince George's County averages 375 car thefts, new study finds

Lede: Prince George's County has reached an average of approximately 375 car thefts per month, new data suggests. The research spans from 2024 to 2026. But this average could be potentially as high as 405 thefts per month and as low as 345 thefts, due to uncertainty in the polling. 

2.  Your editor says "Just tell me if car thefts are going up or down."
    Using only the confidence interval (no significance testing yet),
    what's the most accurate answer you can give? What would you need to
    do next to answer that question more definitively?

The most accurate answer is that while it may look like car thefts have decreased, there is insufficient data to suggest that this change was significant. Next, you would need to find out the data's p-value to determine how significant the decrease was.

When you are finished, save your work, switch to GitHub Desktop, then
add, commit and push your changes to GitHub and submit the URL of the
notebook in ELMS.
