---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "YOUR NAME HERE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
health_inspection_scores <- read_csv("montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
health_inspection_scores |>
  summarise(sd_compliance_score = sd(compliance_score), 
            mean_compliance_score = mean(compliance_score),
            min_compliance_score = min(compliance_score),
            max_compliance_score = max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
health_inspection_scores |>
  ggplot(aes(x = compliance_score)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
   geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed") +
  labs(title = "Distribution of health inspection compliance scores in Montgomery County", x = "Score", y = "# of restaurants")

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**The relatively low standard deviation and the heavily right-skewed histogram tell us that most restaurants comply with Montgomery County health standards, with a large percentage receiving a perfect score. For me, the newsworthy story would be for the the ~1000 restaurants who received an 80 or below. Should restaurant-goers be concerned for such an atypical score, even if it is an 8/10? I'm also curious about the one restaurant that received a 42. **

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
riders <- read_csv ("wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}

wmata_stats <- riders |>
  summarize(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
  )

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- riders |>
  mutate(rail_zscore = (rail - mean(rail)) / sd(rail),
         bus_zscore = (bus - mean(bus)) / sd(bus))
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))
```

***There are 6 unusual days***

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(91)
sample_91 <- riders |> sample_n(91)


sample_91 |> 
  summarize (
  mean_bus_sample = mean(bus),
  sd_bus_sample = sd(bus),
  mean_rail_sample = mean(rail),
  sd_rail_sample = sd(rail)
)
```

```{r}
riders |>
  summarize(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
  )
```

***The results from the random sample are very accurate to the results from the population sample, suggesting that random sampling will often give you an accurate summary of the population sample.***

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

***I think one potentially interesting story lead could be how rails prepare for the holidays. I noticed that 3 of the 6 days were on commonly-celebrated holidays --- Christmas and New Year's Day --- and the ridership was abnormally low. Is business typically difficult on holidays, or are these just anomalies?

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv ("md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |>
  mutate(
    theft_rate_2023 = (thefts_2023/population) * 10000)
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts <- md_car_thefts |>
  mutate(
    median_theft_rate = median(theft_rate_2023),
    total_thefts = sum(thefts_2023)
  )

md_car_thefts |>
  filter(theft_rate_2023 > median_theft_rate) |>
  arrange(desc(theft_rate_2023))
  
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

***What I found most interesting was that St. Mary's and Dorchester were among the top 5 highest theft rates in the state, both well above the median. According to this dataset, they have 2 of the lowest populations of any Maryland county. A potential lede for this could be "Despite being among the least populous counties in Maryland, both Dorchester and St. Mary's counties have some of the highest vehicle theft rates in the state."

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

I would prioritize using z-scores. My editor is looking for outliers, and the best way to find this is to determine which pools deviate the furthest from the mean, which z-scores succintly do.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

As stated earlier, z-scores determine how far away a piece of data deviates from the mean. Because some pools could have a disproportionate number of attendees due to factors outside of its control like lower capacities, this truly determines if a pool's attendance falls outside of what would be generally expected. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

My first concern would be when these 60 days took place. Having the 60 days at the end of the summer could potentially produce lower numbers than the start. What if it rained for weeks straight during this 60-day period? That could also produce lower numbers, especially if dates outside of that were generally sunny. This could potentially become undercoverage bias if some of these factors come into play. 

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
