---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Joe Wicke"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

Nothing immediately sticks out as strange like the Pennsylvania data. If there were binary scores like 100% or 0%, that would be cause for concern. However, I am concerned about how many schools dropped below 20%, not because it's bad, but because it seems so much lower than anything seen in the Pennsylvania data. In fact, every math expectation percentage in this dataset looks generally lower than the scores in the Pennsylvania dataset, and it's strage that none even reach 50%. This might just be indicative of MoCo's education system, but I do wonder if this is truly accurate. 


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

There is a clear negative correlation here. Based on this dataset, it would be safe to assume that there is a significant relationship between the percentage of students on FARMS and proficiency rates. Of course, I doubt this is the only factor, but the relationship certainly exists. 


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

Overall, the model seems much more accurate than the Pennsylvania one. For one, the p-value is less than 0.05, meaning it's statistically significant. The adjusted R-squared tells us that around 2/3rds of the data is accounted for, which is significantly higher than what we saw in the Pennsylvania. While it's still not great, it is technically a majority of the data. It also has a coefficient of about -0.6, meaning it moves significantly in the negative direction. While it's still not perfect, it's definitely more reliable than what we've previously seen. 


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The school with the largest positive residual is Fox Chapel Elementary. The school with the largest negative residual is Germantown elementary.

Fox Chapel Elementary is in Germantown, which has a relatively decent economic profile. IIts median per capita income is a little lower than the median for the state of Maryland, it is higher than the median in the country. Its household income is significantly higher than the national median and slightly higher than the state median. Its racial makeup is pretty diverse, with a 29% Black majority and a close split with its white, Asian, and Hispanic population. 

Strangely, Germantown Elementary is in nearly the same zip code with very similar demographic and income makeups. While there's slight variation with the racial makeup, nothing would suggest any noticeable difference from Fox Chapel Elementary.


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

Maryland data supports this claim -- schools with fewer students on free and reduced-price meals do better on test scores. It's a much different story than Pennsylvania. The data is very clear-cut. But I don't think this is the entire story. If I called for comment, I would also ask about the infrastructure of the schools. How many teachers are at these schools? How satisfied are they with their jobs? How much funding do these schools have? The fact that two elementary schools from the same town have such wildly different results can't be a coincidence. I would start with the two schools in Germantown and compare and contrast how each school operates, then work fromt there. 
