---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

A couple things immediately stand out to me. For one, most of the highest-paying white house staffers are relatively unknown, or at least do not have a large following. The highest, Jacalynne Klopp, doesn't even have an available photo of her online. There are also 8 staffers, including Marco Rubio, who do not make any money working at the White House. This might slightly skew the mean results, though with such a high volume of staffers, its effect should be minimal. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}

wh_salaries |>
summarize(
  mean_salary = mean(SALARY), 
  median_salary = median(SALARY)
)

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
summarize(
  mean_salary = mean(SALARY), 
  median_salary = median(SALARY),
  min_salary = min(SALARY),
  max_salary = max(SALARY)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

As mentioned earlier, the obvious outlier is the 8 staffers who make $0 working for the White House. What I find interesting, however, os that the median salary isn't particularly higher than the U.S. median income. This surprises me becuase I was under the false assumption that White House staffers made millions of dollars, when most make an above-average salary at best. The outlier highest salaries are also complete anomalies, making it puzzling why they make so much more money than the average staffer. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My best story idea is something along the lines of "Here are the 8 White House employees who don't have any salary." I still think that's slightly mind-boggling, especially since someone as high-profile as Rubio is one of the 8. The next step I'd take to develop it is to contact political analysts to see if this is actually something intriguing, or if this is common procedure amongst presidential administrations.
