---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who
are religiously unaffiliated --- a group that makes up about 36% of the
population. In this assignment, you'll run the same simulation for a
different group: White Evangelicals, who make up roughly 10% of the
population.

That difference matters more than it might seem. When a proportion is
close to 50%, samples tend to be pretty well-behaved. When it's much
smaller --- or much larger --- samples can behave differently,
especially at small sizes. Your job is to run the simulation, compare
what you find to the in-class results, and think carefully about what it
means for reporting on religion and politics.

### Task 1: Load the tidyverse

``` {r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class
activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

``` {r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a
respondent's `trad` value is `"White Evangelical"` and `FALSE`
otherwise. Then calculate the proportion of the full dataset that is
White Evangelical and save it as `population_proportion`.

``` {r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- proportion <- mean(ces$evangelical)
```

Write down this value. How does it compare to the unaffiliated
proportion (about 0.36) from the in-class activity?

*The proportion is 0.105, which is lower than the unaffiliated proportion from the in-class activity*

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as
`sample_50`. Calculate the proportion of White Evangelical respondents
in your sample.

``` {r}
set.seed(50)
sample_50 <- ces |> sample_n(50)
proportion2 <- mean(sample_50$evangelical)
```

How close is your sample estimate to the population proportion? How does
the accuracy compare to what you saw with the unaffiliated group in
class?

***The sample estimate is very accurate to the population proportion, which is different from the in-class activity where the sample was not close at all. This tells me that the sample size is much closer to the full dataset's number of White Evangelicals.***

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`.
Calculate the proportion of White Evangelical respondents in this
sample.

``` {r}
set.seed(500)
sample_500 <- ces |> sample_n(500)
proportion3 <- mean(sample_500$evangelical)
```

Did the larger sample improve the estimate? How does the improvement
compare to what you saw in class?

***Unlike in the in-class activity, the estimate was further off when we added more samples to the dataset.***

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but
using your `evangelical` column. Run it as-is.

``` {r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class.
Add a vertical line at `population_proportion`. Make sure to update the
title and subtitle to reflect that this simulation is about White
Evangelicals.

``` {r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: White Evangelicals",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the
unaffiliated histogram from class? Why do you think that is, given that
White Evangelicals make up only about 10% of the population?

***The histogram is less symmetric because there's less data to pull from for white evangelicals than there are for unaffiliated.***

**Compare the n=50 and n=500 histograms for this group.** How much does
the distribution change with a larger sample? Is the improvement bigger
or smaller than what you saw in class with the unaffiliated group?

***The distribution becomes much more narrow with 500 samples, meaning that it is much more accurate, The improvement is slightly more drastic than in the in-class activity.***

**A news organization wants to report on White Evangelical voters using
a poll of 200 people.** Based on your simulations, roughly how many
White Evangelical respondents would that poll be likely to include? What
does that suggest about the reliability of estimates for smaller
subgroups?

**It would likely include 20 White Evangelical respondents. This means that the estimates for smaller subgroups may not accurately represent their wider population, especially compared to larger denominations where you can pull from a larger amount of people.**

**Imagine you are covering the 2028 presidential election and a campaign
shares a poll showing White Evangelical support at 78%.** The poll
sampled 400 people. Based on everything you've seen in this assignment
and in class, what would you want to know before reporting that number?

**Were the amount of White Evangelicals proportional to the population? Was this a randomly selected sample? Did you pull from the same area or a diverse array of places to get this number? 40 responses isn't a lot, so it's key that they are both randomly selected and not all from the same place.***

When finished, save your work, switch to GitHub Desktop, then add,
commit and push your changes to GitHub and submit the URL of the
notebook in ELMS.
