---
title: "final_project"
output: html_document
date: "2026-05-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(purrr)
library(corrplot)
```

***Data exploration: My data covers a wide variety of campus crime from 2016-2023. While the original files had every college in the United States represented, I instead filtered it down to 11 colleges --- the University of Maryland and the ten schools with the largest student bodies in the B1G. Since I do think it would be too broad to just look at general crime trends, I decided to land on burglary, since it's a highly-represented stat across all files. While the data is all-encompassing and easily accessible, the one limitation is a lack of precision in the data. While each year is handily represented, the exact months are not. While it's not impossible to visualize, the data will unfortunately be less detailed in something like a line graph. I initially feared that having 5 separate datasets would make it difficult to analyze the data, I binded the rows together, and thankfully all the data remained. What was nice too was that certain years overlapped in individual datasets, which is now avoided thanks to the combined dataset, which makes everything a little less tricky. For my sakes, anything labeled "BURGLA" will be mainly what I'm looking at. To make my life a little easier, I may make a condensed dataset of just BURGLA and university values. One final issue that may come up is that even in only one campus, many colleges are subdivided by region. I may have to use summarize and sum to get around this (possibly mutate, though that may cause more confusion)***

```{r}
campus_crime_2019 <- read_csv("oncampuscrime2019.csv")
campus_crime_2020 <- read_csv("oncampuscrime2020.csv")
campus_crime_2021 <- read_csv("oncampuscrime2021.csv")
campus_crime_2022 <- read_csv("oncampuscrime2022.csv")
campus_crime_2023 <- read_csv("oncampuscrime2023.csv")
campus_crime_2024 <- read_csv("oncampuscrime2024.csv")

campus_crime_combined <- bind_rows(list(campus_crime_2019, campus_crime_2020, campus_crime_2021, campus_crime_2022, campus_crime_2023, campus_crime_2024))
```
***Data cleaning***

Next, I will isolate burglary and combine campus, so it will all show up as one year:

```{r}
campus_crime_burglary <- campus_crime_combined |>
  select(INSTNM, BRANCH, Total, starts_with("BURGLA")) |>
  pivot_longer(cols = starts_with("BURGLA"), names_to = "year", values_to = "burglary_count") |>
  filter(!is.na(burglary_count)) |>
  mutate(year = as.numeric(str_remove(year, "BURGLA")) + 2000) |>
  group_by(INSTNM, year) |>
  summarize(burglary_count = sum(burglary_count), total_students = max(Total)) |>
  distinct() |>
  ungroup()

##Illinois and Penn State decided to be annoying and duplicate their data. This is to fix it:

campus_crime_burglary_clean <- campus_crime_burglary |>
  mutate(institution_name = case_when(
    INSTNM == "Pennsylvania State University-Main Campus" ~ "Pennsylvania State University",
    INSTNM == "The Pennsylvania State University" ~ "Pennsylvania State University",
    INSTNM == "University of Illinois at Urbana-Champaign" ~ "University of Illinois",
    INSTNM == "University of Illinois Urbana-Champaign" ~ "University of Illinois",
    INSTNM == "Michigan State University" ~ "Michigan State University",
    INSTNM == "Ohio State University-Main Campus" ~ "Ohio State University",
    INSTNM == "Purdue University-Main Campus" ~ "Purdue University",
    INSTNM == "Rutgers University-New Brunswick" ~ "Rutgers University",
    INSTNM == "University of Maryland-College Park" ~ "University of Maryland",
    INSTNM == "University of Michigan-Ann Arbor" ~ "University of Michigan",
    INSTNM == "University of Minnesota-Twin Cities" ~ "University of Minnesota",
    INSTNM == "University of Washington-Seattle Campus" ~ "University of Washington",
    INSTNM == "University of Wisconsin-Madison" ~ "University of Wisconsin"
  ))
```

***Rates***

Now, I will determine the rate per 1000 students for each campus and each year. 

```{r}
campus_crime_burglary_clean |>
  mutate(rate_per_1000 = (burglary_count/total_students) * 1000)
```

Looking at this, my eye immediately gravitates towards the University of Washington's burglary rate in 2020, especially because it's so much higher than any other rate in 2020. The next highest that year hovers around 1.75 per 1000, while UW's is over 4 per 1000. Looking at the University of Maryland's specifically, I initially believe it will show a signifcant amount of decrease in burglaries. It has been consistently going down since 2019, and while I don't believe yet that it will be the lowest, I think it is in a substantially better place compared to other colleges based on initial findings.

***Percent changes***

With that being said, I will now determine each college's percent changes based on their average number of crimes committed per year. I decided to split this into 2 categories: percent change over all years covered and percent change post-pandemic. I've determined "post-pandemic" as 2022 onwards. I was originally also going to do a percent change over the most recent year of the data, but since the data ends at 2023, I merged the two categories. 

First, I'll pivot wide to create a column for every year:

```{r}
burglary_wide <- campus_crime_burglary_clean |>
  select(-total_students, -INSTNM,) |>
   pivot_wider(
    names_from = year,
    values_from = burglary_count,
    values_fn = sum,
    values_fill = 0
  )
```

Then, I will create a column determining the total and average for every university:

```{r}
burglary_wide <- burglary_wide |>
  group_by(institution_name) |>
  mutate(total_burglaries = sum(`2016` + `2017` + `2018` + `2019` + `2020` + `2021` + `2022` + `2023`))

  
burglary_averages <- burglary_wide |>
  mutate(average_burglaries = total_burglaries/8)
  
```

Now, I will figure out each appropriate percent change. I decided to go with 2017 (pre-covid), 2020 (amid covid) and 2023 (post-covid): 

```{r}
burglary_changes <- burglary_averages |>
  mutate(change_2023 = (`2023` - average_burglaries) / average_burglaries * 100) |>
  mutate(change_2020 = (`2020` - average_burglaries) / average_burglaries * 100) |>
  mutate(change_2017 = (`2017` - average_burglaries) / average_burglaries * 100)
  
```

Based on this, Maryland's decrease in crime is somewhere in the middle compared to all other colleges. It currently sits at around 59%, the fifth-highest decrease since 2023. Interestingly, it did have a spike in burglaries around the start of COVID compared to the average, and had the 3rd highest percent increase across all 11 schools. While we will have to find if this decrease is statistically significant, right now it looks like big 10 schools in general are doing better with preventing burglaries, as all 11 decreased from the average in 2023. 


Now, we will perform a t-test to determine if there was a significant difference in pre-covid and post-covid burglary rates. 

Below is the code that I used to conduct the t-test.


```{r}
# Helper: pivot one CSV's BURGLA columns to long format
pivot_burgla <- function(df) {
  df |>
    select(INSTNM, BRANCH, Total, starts_with("BURGLA")) |>
    pivot_longer(cols = starts_with("BURGLA"),
                 names_to = "year",
                 values_to = "burglaries") |>
    mutate(year = as.numeric(str_replace(year, "BURGLA", "")) + 2000) |>
    #I added this to fix duplicates in Penn State and Illinois. I don't mean to brag, but I totally mean to brag.
     mutate(institution_name = case_when(
    INSTNM == "Pennsylvania State University-Main Campus" ~ "Pennsylvania State University",
    INSTNM == "The Pennsylvania State University" ~ "Pennsylvania State University",
    INSTNM == "University of Illinois at Urbana-Champaign" ~ "University of Illinois",
    INSTNM == "University of Illinois Urbana-Champaign" ~ "University of Illinois",
    INSTNM == "Michigan State University" ~ "Michigan State University",
    INSTNM == "Ohio State University-Main Campus" ~ "Ohio State University",
    INSTNM == "Purdue University-Main Campus" ~ "Purdue University",
    INSTNM == "Rutgers University-New Brunswick" ~ "Rutgers University",
    INSTNM == "University of Maryland-College Park" ~ "University of Maryland",
    INSTNM == "University of Michigan-Ann Arbor" ~ "University of Michigan",
    INSTNM == "University of Minnesota-Twin Cities" ~ "University of Minnesota",
    INSTNM == "University of Washington-Seattle Campus" ~ "University of Washington",
    INSTNM == "University of Wisconsin-Madison" ~ "University of Wisconsin"
  ))
    
}
# Pivot each file and combine
all_burglaries <- bind_rows(
  pivot_burgla(campus_crime_2019),
  pivot_burgla(campus_crime_2020),
  pivot_burgla(campus_crime_2021),
  pivot_burgla(campus_crime_2022),
  pivot_burgla(campus_crime_2023),
  pivot_burgla(campus_crime_2024)
) |>
  # Aggregate sub-campuses (e.g., UMD Main + Shady Grove)
  group_by(institution_name, year) |>
  summarize(
    burglaries = sum(burglaries, na.rm = TRUE),
    enrollment = max(Total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Remove duplicate year observations from overlapping CSVs
  distinct(institution_name, year, .keep_all = TRUE)

# Add pre/post COVID label (2016-2019 = pre, 2020-2023 = post)
all_burglaries <- all_burglaries |>
  mutate(period = if_else(year <= 2019, "pre_covid", "post_covid"))

# Compute mean burglary rate per school per period (for paired t-test)
school_period_means <- all_burglaries |>
  group_by(institution_name, period) |>
  summarize(mean_burglaries = mean(burglaries, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = mean_burglaries)

# Paired t-test: each school is a pair (pre vs post)
t_test_result <- t.test(school_period_means$pre_covid, school_period_means$post_covid, paired = TRUE)
```
Now, let's interpret the results of the t-test: 

```{r}
p_value <- t_test_result$p.value
```

Right here, we can see that the p-value comes out to about .063. For our purposes, this would mean that we cannot reject the null hypothesis, which is that there is not a statistically significant change in average pre- and post-covid burglary rates. While the p-value does inch close to the 0.05 threshold, even then, there is no guarantee that a statistically significant change equates to a change that feels significant to an observer. Even if we did deem it statistically significant, the results don't necessarily reflect that. For example, Maryland's mean rate dropped merely by 3 burglaries after COVID, which doesn't feel that significant. This is especially reflected in the confidence interval. Ranging from -1 to 48, it includes zero, which automatically deems it not statistically significant, as the "true" mean difference could be nothing. 

With all that said, there is no significant difference in pre- and post-covid average burglaries across any of the B1G schools listed. 

Next, let's see if higher burglary rates influence higher robbery rates. 

To do this, we'll create a linear regression model:

```{r}
# Modify the pivot helper to grab both crime types
pivot_crimes <- function(df) {
  burgla <- df |>
    select(INSTNM, BRANCH, Total, starts_with("BURGLA")) |>
    pivot_longer(cols = starts_with("BURGLA"), names_to = "year", values_to = "burglaries") |>
    mutate(year = as.numeric(str_replace(year, "BURGLA", "")) + 2000)

  robs <- df |>
    select(INSTNM, BRANCH, starts_with("ROBBE")) |>
    pivot_longer(cols = starts_with("ROBBE"), names_to = "year", values_to = "robberies") |>
    mutate(year = as.numeric(str_replace(year, "ROBBE", "")) + 2000)

  inner_join(burgla, robs, by = c("INSTNM", "BRANCH", "year"))
}

all_crimes <- bind_rows(
  pivot_crimes(campus_crime_2019),
  pivot_crimes(campus_crime_2020),
  pivot_crimes(campus_crime_2021),
  pivot_crimes(campus_crime_2022),
  pivot_crimes(campus_crime_2023),
  pivot_crimes(campus_crime_2024)
) |>
  group_by(INSTNM, year) |>
  summarize(
    burglaries = sum(burglaries, na.rm = TRUE),
    robberies = sum(robberies, na.rm = TRUE),
    enrollment = max(Total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  distinct(INSTNM, year, .keep_all = TRUE)

# Regression: do burglaries predict aggravted ass?
model <- lm(robberies ~ burglaries, data = all_crimes)
summary(model)
```
A couple things imediately stand out to me. For one, there is a positive coefficient of 3, which tells me the regression model is moving significantly. The p-value is also in scientific notation, which tells me that the results are definitely statistically significant. While this definitely leads me to believe that the model is showing a significant positive correlation between burglaries and robberies, one giant red flag is an adjusted R-squared of 0.26, which means that about a quarter of the dataset is represented in this model. Based on this, I think it's safe to say that there is a positive correlation, but the model may be unreliable, as a large chunk of data is unaccounted for. 

Nevertheless, let's visualize it: 

```{r}
all_crimes |>
  ggplot(aes(x = burglaries, y = robberies)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "# of burglaries", y = "# of robberies",
       title = "Do burglaries influence robberies at Big 10 colleges?")
```
Visually, you can see a positive correlation, though the data is slightly more scattered than the results from the model would suggest. Once again, this is where having a higher adjusted R-squared would benefit the data, as there is simply more to draw off of. 
For fun, let's see if we can create a model with a higher adjusted R-squared value. I have two ideas in mind:

First, let's try aggravated assaults:

```{r}
pivot_crimes <- function(df) {
  burgla <- df |>
    select(INSTNM, BRANCH, Total, starts_with("BURGLA")) |>
    pivot_longer(cols = starts_with("BURGLA"), names_to = "year", values_to = "burglaries") |>
    mutate(year = as.numeric(str_replace(year, "BURGLA", "")) + 2000)

  agg_a <- df |>
    select(INSTNM, BRANCH, starts_with("AGG_A")) |>
    pivot_longer(cols = starts_with("AGG_A"), names_to = "year", values_to = "agg_assault") |>
    mutate(year = as.numeric(str_replace(year, "AGG_A", "")) + 2000)

  inner_join(burgla, agg_a, by = c("INSTNM", "BRANCH", "year"))
}

all_crimes_2 <- bind_rows(
  pivot_crimes(campus_crime_2019),
  pivot_crimes(campus_crime_2020),
  pivot_crimes(campus_crime_2021),
  pivot_crimes(campus_crime_2022),
  pivot_crimes(campus_crime_2023),
  pivot_crimes(campus_crime_2024)
) |>
  group_by(INSTNM, year) |>
  summarize(
    burglaries = sum(burglaries, na.rm = TRUE),
    agg_assault = sum(agg_assault, na.rm = TRUE),
    enrollment = max(Total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  distinct(INSTNM, year, .keep_all = TRUE)

model <- lm(agg_assault ~ burglaries, data = all_crimes_2)
summary(model)
```
And to visualize it:

```{r}
all_crimes_2 |>
  ggplot(aes(x = burglaries, y = agg_assault)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "# of aggravated assaults", y = "# of robberies",
       title = "Do aggravated assaults? influence robberies at Big 10 colleges?")
```
Here, there is a much clearer positive correlation. This is supported by a miniscule p-value and a very strong coefficient. However, what concerns me is an adjusted R-squared value of 0.16, which means that most of the data is unaccounted for in this model. While the results are clear with this model, I unfortunately would not call it a reliable model. 

First, let's add some more crime types in our dataset. I'll go with VEHIC, ROBBE, AGGRE, and BURGLA. 

```{r}
pivot_crimes <- function(df) {
  burgla <- df |>
    select(INSTNM, BRANCH, Total, starts_with("BURGLA")) |>
    pivot_longer(cols = starts_with("BURGLA"), names_to = "year", values_to = "burglaries") |>
    mutate(year = as.numeric(str_replace(year, "BURGLA", "")) + 2000)

  agg_a <- df |>
    select(INSTNM, BRANCH, starts_with("AGG_A")) |>
    pivot_longer(cols = starts_with("AGG_A"), names_to = "year", values_to = "agg_assault") |>
    mutate(year = as.numeric(str_replace(year, "AGG_A", "")) + 2000)
  
  robs <- df |>
    select(INSTNM, BRANCH, starts_with("ROBBE")) |>
    pivot_longer(cols = starts_with("ROBBE"), names_to = "year", values_to = "robberies") |>
    mutate(year = as.numeric(str_replace(year, "ROBBE", "")) + 2000)
  
  vehic <- df |>
    select(INSTNM, BRANCH, starts_with("VEHIC")) |>
    pivot_longer(cols = starts_with("VEHIC"), names_to = "year", values_to = "vehicle_thefts") |>
    mutate(year = as.numeric(str_replace(year, "VEHIC", "")) + 2000)

#This I had to look up how to do (I did not use AI):
  list = list(burgla, agg_a, robs, vehic)
  
  crimes <- list |>
     reduce(inner_join, by = c("INSTNM", "BRANCH", "year"))
}

all_crimes_3 <- bind_rows(
  pivot_crimes(campus_crime_2019),
  pivot_crimes(campus_crime_2020),
  pivot_crimes(campus_crime_2021),
  pivot_crimes(campus_crime_2022),
  pivot_crimes(campus_crime_2023),
  pivot_crimes(campus_crime_2024)
) |>
  group_by(INSTNM, year) |>
  summarize(
    burglaries = sum(burglaries, na.rm = TRUE),
    agg_assault = sum(agg_assault, na.rm = TRUE),
    robberies = sum(robberies, na.rm = TRUE),
    vehicle_thefts = sum(vehicle_thefts, na.rm = TRUE),
    enrollment = max(Total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  distinct(INSTNM, year, .keep_all = TRUE)

model <- lm(burglaries ~ agg_assault + vehicle_thefts + enrollment + robberies, data = all_crimes_3)
summary(model)
```
Already, the p-value tells us that this will be statistically significant. The adjust R-squared means about 32% of the dataset is encompassed, which is higher than any other results I was able to get before. Given how large the crime dataset is, I imagine this makes some amount of sense. 

Now, to create the multiple regression model:
```{r}
selected_vars <- all_crimes_3 |>  select(robberies, enrollment, agg_assault, burglaries, vehicle_thefts)
cor(selected_vars)
```
Now to visualize it:

```{r}
corr_matrix <- cor(selected_vars)
corrplot(corr_matrix, method = "circle")
```
From this, we can glean that burglaries and robberies have a significant positive correlaion. Out of all the crimes listed in the dataset, these two easily have the strongest connection. Aggravated assaults and burglaries also have a decently strong correlation, as does aggravated assault and robberies. Interestingly, there's a negative correlation between vehicle thefts and enrollments, though that is a rather weak correlation. Overall, nothing here sounds too surprising. While there are some correlations between crimes, the two explanations are that the crimes either strongly relate (i.e., a burglary can easily turn into a robbery if the burglar confronts the homeowner), or can just be explained by campuses having high crime rates due to the surrounding area. Nothing seems like a major breakthrough when looking at this data. 

***Reporting plan***

Sometimes when looking at datasets, there just isn't an interesting story to be found. Unfortunately, I do think this is the case here. I looked to see if UMD had any abnormal burglary patterns compared to other B1G colleges, and they were relatively insignificant. I looked to see if there were any changes in burglary trends post-pandemic, and the changes were insignificant. I looked to see if burglary correlates to any other crime, and the results did not produce anything particularly newsworthy. However, I think I have two leads I could potentially follow. 

While post-covid burglary trends did not garner a statistical significance, I wonder if that's an interesting story to follow. After the pandemic, there had to have been some plan in place to prevent crime as it inevitably rises. What have B1G colleges been saying about their crime rates? Are they satisfied with its progress, or do they want to continue improvements? Do the statements from the universities contradict the (lack of) statistical significance in the data? I think there could be an interesting story, especially if the campus police go against what the data says. 

For this, I would need to contact campus police officers, maybe a police chief, a statement or two from some B1G university presidents. Perhaps some interviews from students and local residents can add some perspective. 

Building off this a little, I'm also interested in the University of Washington's burglary rates pre-and post-pandemic. They are the only university to increase in burglaries following the pandemic, and had the highest change from average crimes during the pandemic. I think there's potentially an interesting story here. Why do they have the most amount of burglaries despite only being the 4th most-populous campus? Is local law enforcement at all concerned? I think I would talk to all the same sources I would for the first pitch idea. I think even a combination of the two can work to some capacity. 

Of course, the one caveat for both of these stories is that burglary is only one subsect of crime, and does not solely contribute to a campus's overall safety. Regardless, I think it could play an interesting part in a larger conversation about safety at B1G schools. 
